---
title: "Darwin Core mapping"
subtitle: "For: monitoring of LIFE RIPARIAS invasive alien plants in the Flemish part of the LIFE RIPARIAS areas"
author:
- Damiano Oldoni
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

# Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE)
```

Install required libraries (only if the libraries have not been installed before):

```{r}
installed <- rownames(installed.packages())
required <- c("tidyverse",
              "tidylog",
              "here",
              "readxl",
              "sf",
              "DBI",
              "RSQLite",
              "digest",
              "testthat"
)
if (!all(required %in% installed)) {
  install.packages(required[!required %in% installed])
}
```

Load libraries:

```{r message = FALSE}
library(tidyverse)      # To do data science
library(tidylog)        # To provide feedback on dplyr functions
library(here)           # To find files
library(readxl)         # To read Excel files
library(sf)             # To convert coordinate systems
library(DBI)            # To work with databases
library(RSQLite)        # To work with SQLite databases in R
library(digest)         # To create hashes
```

# Read source data

Create a data frame `occurrences` from the source data:

```{r read_data}
occurrences <- readxl::read_excel(
  path = here("data", "raw", "Plantenmonitoring RIPARIAS (Vlaanderen).xlsx"),
  sheet = "Data_inventarisaties_opgekuist")
```

Preview data:

```{r}
occurrences %>% head(n = 10)
```

# Process source data

## Add scientific name where missing

Column `soort` contains vernacular names in Dutch, sometimes with the scientific name as prefix: 

```{r soort}
occurrences %>%
  distinct(soort)
```
We add the scientific name where missing:
- `roze waterlelie`: _Nymphaea marliacea_
- `brede waterpest`: _Elodea canadensis_
- `smalle waterpest`: _Elodea nuttallii_

```{r add_Nymphaea_marliacea}
occurrences <-
  occurrences %>%
  mutate(soort = recode(
    soort,
    "roze waterlelie" = "roze waterlelie / Nymphaea marliacea",
    "brede waterpest" = "brede waterpest / Elodea canadensis",
    "smalle waterpest" = "smalle waterpest / Elodea nuttallii"
  )
)
```

Show distinct values of column `soort`:

```{r show_soort_after_mapping}
occurrences %>% 
  distinct(soort)
```

## Tidy data

Extract vernacular names from column `soort`:

```{r vernacular}
occurrences <- 
  occurrences %>%
  mutate(
    vernacularName = stringr::str_extract(
      string = .data$soort,
      pattern = "(.*?)(?=\\/)")) %>%
  mutate(vernacularName = stringr::str_trim(.data$vernacularName))
```

Extract scientific names from column `soort`:

```{r scientific}
occurrences <- 
  occurrences %>%
  mutate(
    scientificName = stringr::str_extract(
      string = .data$soort,
      pattern = "(?<=\\/)(.*)")) %>%
  mutate(scientificName = stringr::str_trim(scientificName))
```

Mapped values:

```{r sc_vernacular}
occurrences %>% distinct(scientificName, vernacularName)
```

# Create database

Create a SQLite database with the source data, so it can be queried with SQL in the next steps:

```{r create_db}
message("Create in memory SQLite database...")
con <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
# import occurrences
DBI::dbWriteTable(con, "occurrences", occurrences)
message("DONE")
```


# Darwin Core mapping

Create [Event](https://rs.gbif.org/core/dwc_event_2022-02-02.xml) core:

```{r event}
dwc_event_sql <- glue::glue_sql(
  readr::read_file(here::here("sql", "dwc_event.sql")),
  .con = con
)
dwc_event <- DBI::dbGetQuery(con, dwc_event_sql)
```

Create [Occurrence](https://rs.gbif.org/core/dwc_occurrence_2022-02-02.xml) extension:

```{r occurrence}
message("Map occurrences to DwC...")
dwc_occurrence_sql <- glue::glue_sql(
  readr::read_file(here::here("sql", "dwc_occurrence.sql")), 
  .con = con
)
dwc_occurrence <- DBI::dbGetQuery(con, dwc_occurrence_sql)
message("DONE")
```

# Save data to CSV

```{r save_csv}
write_csv(dwc_event, here::here("data", "processed", "event.csv"), na = "")
write_csv(
  dwc_occurrence, here::here("data", "processed", "occurrence.csv"),
  na = ""
)
```
